Radiocarbon measurement of CO~2~ from carbonate minerals using a gas ion source with an open split interface

### 2022-04-03 doc/AMS-15/AMS-15_paper/AMS-15_paper.Rmd
Complete reference standards section. Lots of edits.

---
title: "Radiocarbon measurement of CO~2~ from carbonate minerals using a gas ion source with an open split interface"
author:
- Brett Longworth
- Josh Burton
- Simon Pendleton
- Sydney Moser
- Mark Roberts
- Mark Kurz
output:
  word_document: default
  pdf_document: default
  html_document: default
bibliography: HGIS_paper.bib
csl: elsevier-without-titles.csl
---

@@

The gas-capable 134-sample MCSNICS from National Electrostatics Corporation (NEC) [@Xu2007\] works on the same principle as hybrid sources developed by ETH Zürich \[@ruff_gas_2007; @Ruff2010] and High Voltage Engineering Europa (HVEE) [@BronkRamsey2004]. The sample CO~2~ - helium mixture is introduced onto a titanium insert or frit in an aluminum target cathode using stainless tubing mounted to an arm with a probe that mates with the back of the cathode. Sample gas is exposed to a beam of charged cesium atoms, which causes dissociation of CO~2~ and ionization of molecular and atomic species. Negative ions are extracted from the target, carbon species are selected, and interfering masses filtered out by the USAMS Tandetron beamline.

@@

Reference and process blank materials were not pretreated or cleaned. Carbonate shell samples for the core comparison study were picked from a split core and cleaned by washing with distilled water and dried at 60 C. Shell samples were crushed and mixed, but not otherwise pretreated. Other? Sonication?

Standards were prepared by evacuating vials and then filling with a modern standard CO~2~ using the automated gas handling system. Standards were adjusted to the same size as other samples by removing CO~2~ and replacing with helium using a syringe. Process blank samples were prepared from powdered C-1 or TIRI-F, acidified in vials using the method above.

@@

Measurements were corrected for process and measurement blank using a linear correction that assumes that the corrected fraction modern of blank samples should be zero, samples with the fraction modern of the standard need no correction, and applies a proportional correction to samples of intermediate fraction modern \[@roberts_14c_2019\]. The correction uses measured values of standard gas and blank samples for correction of contaminant carbon and machine blank. The errors in the blanks and standards were propagated with errors of the samples to produce the final error estimate for the sample. This correction does not account for the effect of a blank with constant mass or current on samples of varying mass or current. For all measurements used to assess the performance of the method and for the Haiti core comparison, sample, standard, and blank masses were kept constant. Sample current did vary, but we believe the linear correction described above produces the most accurate results within this range. We discuss this constant contaminant blank in the results below. Data were reduced using a package written in R [@r_core_team_r_2022] which is published on github (<https://github.com/blongworth/hgis>).

@@

Initial work with the NEC source established optimal source parameters for consistent source performance. We used tank CO~2~ connected through a regulator and glass capillary to the ion source to measure source performance over a range of CO~2~ flows. Best currents for the source was 10 μA ^12^C^-^, or roughly 1/10 of graphite target current. Peak current was produced at a CO~2~ flow rate of 7.5 μl min^-1^ CO~2~, but the source produced near optimal currents with flows from 7.5-25 μl min^-1^ CO~2~. This is comparable with other hybrid sources [@Xu2007], although improved target design and continued optimization of source parameters may improve performance \[@Salazar2013; \@Fahrni2013\]. Typical currents for this study were between 4-9 μA and average current for all samples was `r curmean`. Several studies have shown best currents at lower CO~2~ flow rates [@Xu2007; @Ruff2010]. While efficiency of the source is best at very low flow rates (\<1.5 μl min^-1^), measuring samples with flow set for optimal current produces more precise results due to higher count rates.

@@

Precision of measurements of modern CO~2~ has been better than 1% (std error of mean of measurements of a target) for high current samples, with precision dropping off as current decreases below 6 μA or so. Accuracy is reasonable within the large errors on gas measurements. Uncorrected measurements of dead CO~2~ range from 1% to 5% of modern depending on sample current.

@@

Where $C(t)$ is the fraction CO~2~, $r$ is the rate of helium displacement flow, and $V$ is the initial volume of the CO~2~-filled vial. Figure 2c shows the resulting relationship of sample ion current to CO~2~ flow rate. The most important feature of the current-flow relationship is the wide range of flows under which stable ion current is measured, which in turn produces relatively stable currents and ratios despite a changing carbon flow rate during measurement.

@@

The HGIS method produces higher ratios for radiocarbon dead samples than the graphite method. Measurement of radiocarbon-free reference materials shows that the average normalized ratio for these samples is `r blankmean` ± `r blanksd` (N=`r nblank`). We investigated the sources of this method blank as a current-dependent blank and as a current-independent blank.

HGIS cathodes produce measurable ion current, even with no gas flow. This current is initially \> 10 μA and drops exponentially, reaching a baseline of \~100 nA after 5-10 minutes. For the range of currents in this study (3-9 μA), this baseline current represents a 1-6% contribution to the total measured sample current at the F^14^C of the target current baseline. This can be interpreted as a constant current dependent blank, analogous to a mass-dependent "constant contaminant" blank and may be modeled with the same tools [@haghipour_compound-specific_2019; @hanke_comprehensive_2017; @roberts_14c_2019; @sun_14_2020\]. To assess blank contribution and ratio dependence on sample current, vials of live and dead standard gas were measured as the displacement helium flow diluted the samples, producing diminishing current over time. We used these data to develop a current-balance model following methods used to assess mass-dependent or constant contamination blanks in graphite targets \[@shah_walter_ultra-small_2015; \@sun_14_2020]. The constant contamination blank was also assessed directly by measuring pure helium in vials using the same method. Results from both tests are shown in Figure 3a. The inverse of the sample current is used to linearize the data, allowing assessment with a simple linear model. Fits from modern and dead dilution tests should trend towards the fraction modern of the blank, crossing at the inverse of blank current. Based on this assessment, we estimate the ratio of the blank as having F^14^C of 0.43 with a current of 110 nA.

@@

Precision and accuracy of the method were assessed by measuring carbonate standards with consensus values from graphite measurements. Summary data for these samples are listed in Table 1, and measurement data for all samples is in supplementary information. Measurement, data reduction, and blank correction was performed as described above.

Overall, reference materials run with the hybrid gas method were in good agreement with consensus values. Mean difference from consensus for all measurements was -0.04 pMC, and the standard deviation of the difference was 0.011 F^14^C. Mean propagated measurement error was 0.011 F^14^C for these samples. The hybrid method has no bias in aggregate. Method error is larger than graphite, but calculated per-sample error is similar to standard deviation of replicate measurements. A summary of accuracy and precision of these measurements is in Table 1.

The difference between reference samples and graphite consensus does not vary systematically by fraction modern, indicating that the method has no significant bias by fraction modern (Figure 4).

The per-sample error agrees well with variability of replicates, and measurement of known-value standards shows that the method has no significant bias.

@@

age range

We found the HGIS and graphite measurements to be in good agreement, with no significant bias (Figure 5a). Mean difference between hybrid and graphite measurements was -0.003 F^14^C and the standard deviation of this difference was 0.009 F^14^C. Mean per-sample error was 0.010 F^14^C for HGIS measurements, compared with 0.002 F^14^C for graphite.

@@

We applied a correction for the regional marine reservoir effect with a delta R of -216 and uncertainty 92 based on the oxcal model with 10 reservoir points. Samples with radiocarbon age \> 0 were calibrated using the Marine20 curve @heaton_marine20marine_2020. The NH1 curve from Hua et. al @hua_atmospheric_2013 was used for calibration of postmodern dates.

@@

We developed a method for the analysis of carbonates using an open-split interface to a hybrid sputter gas ion source. This system produces results that agree with graphite measurements to better than 1% from 8-20 mg of carbonate. Sample preparation takes about 10 min. of person time and 6 h of automated processing and reaction time, and 20 min. of measurement time per sample. The simplicity of of this approach reduces costs, improves reliablity, and reduces system maintenance. With lower cost and time per sample, the lower precision of the method may be balanced by the ability to measure more samples from a population, i.e. triplicate measurement or more core depths. This may lead to more accurate results compared to a more precise graphite measurement where reproducibility is not as well constrained.

@@

This work was supported by the NOSAMS NSF cooperative agreement 1755125. We are grateful to C. McIntyre, R. Hansman, G. Salazar, T. Ognibene, and T. Broek for discussions on gas handling and gas ion sourcery; and to J. Gunnard for improvements to the text.\

@@

Figure 2. Performance of the NOSAMS MCSNICS hybrid gas ion source. a. Modeled flows of helium and CO~2~ to the ion source during displacement of a pure CO~2~ vial sample with helium. b. Measured 12C- ion current during the same dilution experiment. c. Measured ion current for modeled CO~2~ flow. The stability of the source over a range of flow produces stable current for up to 40 m of measurement time.

Will probably skip this figure: Figure 3. Relationship of sample size to ^12^C^-^ ion current. There is no strong relationship between size and current for samples larger than 800 μg C.

@@

Figure 5. Comparison of Haiti core samples measured as graphite and via HGIS. a. Difference between HGIS and graphite measurements of the same materials. Difference between HGIS and graphite measurements is shown on the vertical axis and plotted against the mean of the two measurements. Samples are offset on the horizontal axis for clarity. Error bars are measurement error for each sample. Graphite measurements are included to allow comparison of error estimates. b) comparison of age-depth models produced using HGIS and graphite measurements.

>>>>

---
title: "Radiocarbon measurement of CO~2~ from carbonate minerals using a gas ion source with an open split interface"
author:
- Brett Longworth
- Josh Burton
- Simon Pendleton
- Sydney Moser
- Mark Roberts
- Mark Kurz
output:
  html_document: default
  pdf_document: default
  word_document: default
bibliography: HGIS_paper.bib
csl: elsevier-without-titles.csl
---

@@

The gas-capable 134-sample MCSNICS from National Electrostatics Corporation (NEC) [@Xu2007] works on the same principle as hybrid sources developed by ETH Zürich [@ruff_gas_2007; @Ruff2010] and High Voltage Engineering Europa (HVEE) [@BronkRamsey2004]. The sample CO~2~ - helium mixture is introduced onto a titanium insert or frit in an aluminum target cathode using stainless tubing mounted to an arm with a probe that mates with the back of the cathode. Sample gas is exposed to a beam of charged cesium atoms, which causes dissociation of CO~2~ and ionization of molecular and atomic species. Negative ions are extracted from the target, carbon species are selected, and interfering masses filtered out by the USAMS Tandetron beamline.

@@

Reference and process blank materials were not pretreated. Carbonate shell samples for the core comparison study were picked from a split core and cleaned by washing with distilled water and dried at 60 C. Shell samples were crushed and mixed, but not otherwise pretreated.

Standards were prepared by evacuating vials and then filling with a modern standard CO~2~ using the automated gas handling system. Standards were adjusted to the same size as other samples by extracting an appropriate volume of CO~2~ from the full vial and replacing with helium using a syringe. Process blank samples were prepared from powdered C-1 or TIRI-F, acidified in vials using the method above.

@@

Measurements were corrected for process and measurement blank using a linear correction that assumes that the corrected fraction modern of blank samples should be zero, samples with the fraction modern of the standard need no correction, and applies a proportional correction to samples of intermediate fraction modern $$@roberts_14c_2019$$. The correction uses measured values of standard gas and blank samples for correction of contaminant carbon and machine blank. The errors in the blanks and standards were propagated with errors of the samples to produce the final error estimate for the sample. This correction does not account for the effect of a blank with constant mass or current on samples of varying mass or current. For all measurements used to assess the performance of the method and for the Haiti core comparison, sample, standard, and blank masses were kept constant. Sample current did vary, but we believe the linear correction described above produces the most accurate results within this range. We discuss this constant contaminant blank in the results below. Data were reduced using a package written in R [@r_core_team_r_2022] which is published on github (<https://github.com/blongworth/hgis>).

@@

Initial work with the NEC source established optimal source parameters for consistent source performance. We used tank CO~2~ connected through a regulator and glass capillary to the ion source to measure source performance over a range of CO~2~ flows. Best currents for the source was 10 μA ^12^C^-^, or roughly 1/10 of graphite target current. Peak current was produced at a CO~2~ flow rate of 7.5 μl min^-1^ CO~2~, but the source produced near optimal currents with flows from 7.5-25 μl min^-1^ CO~2~. This is comparable with other hybrid sources [@Xu2007], although improved target design and continued optimization of source parameters may improve performance $$@Salazar2013; \@Fahrni2013$$. Typical currents for this study were between 4-9 μA and average current for all samples was `r curmean`. Several studies have shown best currents at lower CO~2~ flow rates [@Xu2007; @Ruff2010]. While efficiency of the source is best at very low flow rates (\<1.5 μl min^-1^), measuring samples with flow set for optimal current produces more precise results due to higher count rates.

@@

@@

Where $C(t)$ is the fraction CO~2~, $r$ is the rate of helium displacement flow, and $V$ is the initial volume of the CO~2~-filled vial. Figure 2c shows the resulting relationship of sample ion current to CO~2~ flow rate. The most important feature of the current-flow relationship is the wide range of flows under which stable ion current is measured, which in turn produces relatively stable currents and ratios despite a changing carbon flow rate during measurement. Several studies have shown peak currents over a narrower range of flows @Fahrni2013; @Broek2022 at high helium dilution rates (95%), while Xu et. al @Xu2007 shows a somewhat broader peak with pure CO~2~. Broek et. al @Broek2022 shows peak broadening with decreasing dilution ratio. While lower rates of dilution decrease current slightly, the observed peak broadening allows stable measurement over the range of CO~2~ flows produced by the open split.

@@

The HGIS method produces higher ratios for radiocarbon dead samples than the graphite method. Measurements of radiocarbon-free reference materials (IAEA C1 marble and TIRI F spar calcite) show that the average normalized ratio for these samples is `r blankmean` ± `r blanksd` (1σ uncertainty, N=`r nblank`). We investigated the sources of this method blank as a current-dependent blank and as a current-independent blank, and implemented a blank correction for the method.

HGIS cathodes produce measurable ion current, even with no gas flow. This current is initially \> 10 μA and drops exponentially, reaching a baseline of \~100 nA after 5-10 minutes. For the range of currents in this study (3-9 μA), this baseline current represents a 1-6% contribution to the total measured sample current at the F^14^C of the target current baseline. This can be interpreted as a constant current dependent blank, analogous to a mass-dependent "constant contaminant" blank and may be modeled with the same tools [@haghipour_compound-specific_2019; @hanke_comprehensive_2017; @roberts_14c_2019; @sun_14_2020]. To assess blank contribution and ratio dependence on sample current, vials of live and dead standard gas were measured as the displacement helium flow diluted the samples, producing diminishing current over time. We used these data to develop a current-balance model following methods used to assess mass-dependent or constant contamination blanks in graphite targets [@shah_walter_ultra-small_2015; \@sun_14_2020]. The constant contamination blank was also assessed directly by measuring pure helium in vials using the same method. Results from both tests are shown in Figure 3a. The inverse of the sample current is used to linearize the data, allowing assessment with a simple linear model. Fits from modern and dead dilution tests should trend towards the fraction modern of the blank, crossing at the inverse of blank current. Based on this assessment, we estimate the ratio of the blank as having F^14^C of 0.43 with a current of 110 nA.

@@

```{r, message = FALSE}
#library(gt)
library(errors)

options(digits = 3)
options(scipen = 4)
options(decimals = 4)
options(errors.notation = "plus-minus")

# Format data
cons_df <- carb %>% 
  filter(!(wheel %in% c("USAMS020521", "USAMS061121", "USAMS061821"))) %>% 
  mutate(he12C = he12C * 1E6) %>% 
  compare_consensus() %>% 
  filter(Name %in% c("TIRI-I", "C-2", "NOSAMS2"))

# Summary table
cons_df %>% 
  group_by(Name, fm_consensus) %>% 
  summarize(across(c(fm_corr, sig_fm_corr, fm_diff, sigma),
                   list(mean = mean, sd = sd)),
            N = n()) %>% 
  ungroup() %>% 
  arrange(fm_consensus) %>% 
  mutate (fm_corr_err = set_errors(fm_corr_mean, fm_corr_sd),
          Name = case_when(Name == "TIRI-I" ~ "TIRI I",
                           Name == "C-2" ~ "IAEA C2",
                           Name == "NOSAMS2" ~ "NOSAMS 2")) %>% 
  select(`Sample Name` = Name, 
         `Reference F^14^C` = fm_consensus,
         `HGIS F^14^C ± stdev` = fm_corr_err,
         `Mean per-sample error` = sig_fm_corr_mean,
         N) %>% 
  knitr::kable(digits = 3,
               caption = "Reference material precision and accuracy. Reference 
               values are community consensus values for TIRI I and C2, and the 
               NOSAMS internal graphite value for NOSAMS2. NOSAMS2 is a 
               homogenized bivalve molusk reference. The mean of per-sample 
               error estimates is included for comparison to the reproducibilty 
               of the samples.")

  # gt() %>% 
  # tab_header(title = "Reference material precision and accuracy",
  #            subtitle = "Values in pMC, Errors are SD of measurements") %>% 
  # fmt_number(4, 
  #            decimals = 3) %>% 
  # cols_align("right", 3)

# Consensus summary for inline data
cons_sum <- cons_df %>% 
  filter(Name %in% c("TIRI-I", "C-2", "NOSAMS2")) %>% 
  ungroup() %>% 
  summarize(across(c(sig_fm_corr, fm_diff),
                   list(mean = mean, sd = sd)),
            N = n()) 
```

Precision and accuracy of the method were assessed by measuring carbonate reference standards for comparison with consensus values from graphite measurements. We measured TIRI I, a roughly quarter-modern turbidite carbonate, IAEA C2, a half modern turbidite carbonate, and NOSAMS 2, a modern internal shell carbonate standard from a bivalve molusk REF. Summary data for these samples are given in Table 1, and measurement data for all samples is in supplementary information. Measurement, data reduction, and blank correction was performed as described above.

Reference materials run with the hybrid gas method were in good agreement with consensus values. Mean difference from consensus for all measurements was F^14^C `r format(cons_sum$fm_diff_mean, digits =1, nsmall = 4)` shows that the hybrid method has no bias in aggregate. The standard deviation of the difference in F^14^C was `r format(cons_sum$fm_diff_sd, nsmall = 4)`, which may be taken as an estimate of the 1σ precision of the method. The mean of propagated measurement errors for these samples was `r format(cons_sum$sig_fm_corr_mean, nsmall = 4)` F^14^C for these samples, which is in good agreement with the method precision above. The method error is about three times larger than the roughly 3 ‰ precision of graphite measurements, but is within the range of method precision of other gas sources REFS.

The difference between reference samples and graphite consensus does not vary systematically by F^14^C, indicating that the method has no significant bias with radiocarbon content (Figure 4). Each reference standard type agrees with graphite consensus values within the standard deviation of samples measured, with slightly more measurements agreeing with graphite within 1σ uncertainties than would be expected statistically.

@@

We found the HGIS and graphite measurements to be in good agreement, with no significant bias over the range in F^14^C of the core samples (0.4 - 1.15) (Figure 5a). Mean difference between hybrid and graphite measurements was -0.003 F^14^C and the standard deviation of this difference was 0.009 F^14^C. Mean per-sample 1σ uncertainty was 0.010 F^14^C for HGIS measurements, compared with 0.002 F^14^C for graphite. There is an apparent, but not significant positive relationship between the HGIS-graphite difference and measured value. The higher graphite values at lower F^14^C may indicate an excessive HGIS blank correction for these samples or an undercorrected method blank for graphite measurements.

@@

We applied a correction for the regional marine reservoir effect with a delta R of -216 and uncertainty 92 based on the oxcal model with 10 reservoir points. Samples with pre-bomb radiocarbon ages were calibrated using the Marine20 curve @heaton_marine20marine_2020. The NH1 curve from Hua et. al @hua_atmospheric_2013 was used for calibration of postmodern dates, but these measurements were not included in the age model due to uncertainties in reservoir correction.

@@

We developed a method for the analysis of carbonates using an open-split interface to a hybrid sputter gas ion source. This system produces results that agree with graphite measurements to better than 1% from 8-20 mg of carbonate. Sample preparation takes about 10 min. of person time and 6 h of automated processing and reaction time, and 20 min. of measurement time per sample. The simplicity of of this approach reduces costs, improves reliability, and reduces system maintenance. With lower cost and time per sample, the lower precision of the method may be balanced by the ability to measure more samples from a population, i.e. triplicate measurement or more core depths. This may lead to more accurate results compared to a more precise graphite measurement where reproducibility is not as well constrained.

@@

This work was supported by the NOSAMS NSF cooperative agreement 1755125. We are grateful to C. McIntyre, R. Hansman, G. Salazar, T. Ognibene, and T. Broek for discussions on gas handling and gas ion sourcery; and to J. Gunnard for improvements to the text.

@@

Figure 2. Performance of the NOSAMS MCSNICS hybrid gas ion source. (a) Modeled flows of helium and CO~2~ to the ion source during displacement of a pure CO~2~ vial sample with helium. (b) Measured 12C- ion current during the same dilution experiment. (c) Measured ion current for modeled CO~2~ flow. The stability of the source over a range of flow produces stable current for up to 40 m of measurement time.

@@

Figure 5. Comparison of Haiti core samples measured as graphite and via HGIS. (a) Difference between HGIS and graphite measurements of the same materials. Difference between HGIS and graphite measurements is shown on the vertical axis and plotted against the mean of the two measurements. Samples are offset on the horizontal axis for clarity. Error bars are measurement error for each sample. Graphite measurements are included to allow comparison of error estimates. (b) Comparison of age-depth models produced using HGIS and graphite measurements.
